Prospects of detecting gamma-ray emission from galaxy clusters: cosmic rays and dark matter annihilations

### 2011-05-07 paper.tex
corrected text, minor restructuring

\author{Anders Pinzke$^{1}$} \email{apinzke@physics.ucsb.edu} 
\author{Christoph Pfrommer$^{2}$}\email{pfrommer@cita.utoronto.ca}
\author{Lars Bergstr\"om$^{3}$}\email{lbe@fysik.su.se}

@@

\begin{abstract}
... to be written...  if $\mlim \approx 10^{-6}\,\msun$ then the boost
  from SFE is smaller than about a factor 5 in the MW and 2 in Fornax, or
  alternatively if SFE corrext then $\mlim > 10^2\,\msun$.
\end{abstract}

@@

3. Galaxy clusters. This possibility has been less studied, however we
noted in a previous Letter \cite{2009PhRvL.103r1302P} that the are
certain advantages that work in favour of this possible target for
gamma-ray detection of dark matter annihilation.  Galaxy clusters
constitute the most massive objects in our Universe that are forming
today. This causes their DM subhalo mass function to be less affected
by tidal stripping compared to galaxy sized halos that formed long
ago.  The annihilation luminosities of the DM halo component for
e.g. the Virgo cluster and the Draco dwarf scales in a way (see
\cite{2009PhRvL.103r1302P}) that the ratio of gamma-ray luminosities
from the smooth components is around 4, in favour of Virgo. In
addition, there may be a further enhancement due to substructure,
which to a large extent should be unaffected by tidal disruption, at
least in the outer regions.  According to a recent estimate
\cite{2011MNRAS.410.2309G}, more massive haloes tend to have a larger
mass fraction in subhalos.  For example, cluster size haloes typically
have 7.5 per cent of the mass within $r_{200}$ in substructures of
fractional mass larger than $10^{-5}$, which is 25 per cent higher
than galactic haloes \footnote{All halo masses and length scales in
  this paper are scaled to the currently favored value of Hubble's
  constant, $H_0 = 70\, \rmn{km~s}^{-1}\,\rmn{Mpc}^{-1}$. We define the
  virial mass $\mvir$ and virial radius $\rvir$ as the mass and radius
  of a sphere enclosing a mean density that is 200 times the critical
  density of the Universe $\rho_{\rmn{cr}}$.}.

@@

This yields Fornax ($\mvir=10^{14}\,\msun$) and Virgo
($\mvir=2.1\times10^{14}\,\msun$) \del{\cite{}} as the prime targets for DM
observations, Perseus ($\mvir=7.7\times10^{14}\,\msun$) for CR induced
emission and we additionally decide in favor of the well studied
cluster Coma ($\mvir=1.4\times10^{15}\,\msun$) for comparison.

@@

This work studies the prospects for detecting extended gamma-ray
emission from galaxy clusters induced by both DM as well as CR
protons. In section two we discuss the theory of DM and gamma-rays. In
particular, we focus one the leptophilic and supersymmetric DM models
used in this work, and outline the framework for estimating the
gamma-ray emission from various radiative processes. In section three
we calculate the gamma-ray fluxes and spectral distributions from four
clusters and compare it to upper limits set by Fermi-LAT. For the same
clusters we derive the gamma-ray surface brightness profiles in
section four. To extend the analysis to more clusters and find
promesing targets we estimate the fluxes from DM and CRs of all
clusters in the extended HIFLUGCS sample in section five. In section
six we conclude and discuss our results.

% --- section:  --- %
\section{Theory}
\label{sect:theory}
We start the section by discussing two different, but well motivated DM
models; leptophilic and supersymmetric DM. In addition we present the
gamma-ray framework that is used to calculate the emission for these DM
models using an Einasto DM density profile and the expected
enhancement from DM substructures. In the end of this sections we
outline the framework of inverse Compton emission on both CMB photons
as well as photons from stars and dust.

\subsection{Detecting Particle Dark Matter}
\label{sect:PF}
Besides the intrinsic interest in the gamma-ray flux from galaxy
clusters generated by conventional hadronic and electromagnetic
process - which should be close to observability with the Fermi-LAT
data \cite{1997ApJ...487..529B,2007A&A...473...41E,2010MNRAS.409..449P}
- the possible contribution from WIMP dark matter is of greatest
interest. A WIMP (Weakly Interactive Massive Particle) which fulfils
the WMAP bounds on the relic density of cold dark matter (CDM) of
\cite{Komatsu:2010fb}
$$\Omega_{CDM}h^2=0.112\pm 0.0056,$$ will in many cases naturally give
a $\gamma$ yield which may be observable. In addition, there are
possible enhancement effects known, such as the astrophysical boost
from dense substructure of dark matter halos, or the particle physics
boost from the Sommerfeld effect, which may increase the chances of
detection further.

There are three methods for detecting dark matter candidates which are
presently employed: First, at particle physics accelerators such as
CERN's LHC, one is now entering into a new energy regime which may
allow the production of the heavy, electrically neutral and long-lived
particles which may constitute dark matter. From these experiments one
may get the first glimpse of the mass scale beyond the standard model
where dark matter may reside. However, to really show that any of the
hypothetical, new particles created at the LHC is actually the dark
matter, one has to rely on the two other methods available for the
dark matter search, namely direct and indirect detection. Direct
detection methods, which are presently evolving rapidly use the feeble
interaction between that dark matter particles of the galactic halo
and nuclei such as Germanium, Sodium, Iodine, Argon or Xenon to infer
the scattering cross section and a rough estimate of the mass of the
dark matter particle (for the currently most sensitive search, see
\cite{Aprile:2010um}). A characteristic of this method is that it
basically only depends on the local dark matter density (which is
rather well determined to be around $0.4$ GeV/cm$^3$) and the basic
cross section for dark matter - nucleus scattering. A disadvantage is
that it does benefit from the two enhancement mechanisms mentioned
above, boost from substructure and/or the Sommerfeld effect. Also, it
cannot be excluded - although it seems at present improbable - that
the local halo structure is such that the solar system happens to be
in an underdense region.

@@

There is also a possibility to search for dark matter annihilation
indirectly in antimatter channels, like antiprotons or positrons. Here
diffusion is, however, very efficient which means that the directional
signature which is so important for gamma-rays, is completely
erased. This method has, however, recently been much in focus due to
the surprising rise with energy of the electron to proton ratio
measured by the PAMELA satellite experiment \cite{Adriani:2008zr} and
the ``bump'' in the total electron plus positron spectrum detected by
Fermi \cite{Abdo:2009zk}. The viable particle physics models
describing these results contain some non-trivial, or even unnatural
elements, however. First of all, annihilation to quarks has to be very
much suppressed, due to the non-observation of an enhanced antiproton
flux by PAMELA \cite{Adriani:2010rc}. This means that only so-called
{\em leptophilic} models, where the annihilation only goes to leptons
have to be considered. Here, one has to have a sizeable fraction going
to muons and tau leptons, as the shape of the spectrum disfavours
direct annihilation into electrons and positrons. To get agreement
with the measured absolute rates, one also has to have a large boost
factor (of the order of at least several hundred), which is difficult
to achieve with substructure alone. An additional enhancement of the
annihilation rate has to be invoked, for example given by the {\em
  Sommerfeld effect}. This basically means that the usual connection
between annihilation rate at decoupling (when relative velocities $v$
were of order $0.2 - 0.3$), which gives the relic density, and
present-day annihilation in the halo (with velocities $v\sim 10^{-3}$)
is broken. This would come about because the dark matter sector
contains additional light particles which mediates an interaction with
almost bound states which enhance the wave function at the relative
origin, and therefore the annihilation rate (for a pedagogical review
of viable scenarios, see \cite{ArkaniHamed:2008qn}). As a final
difficulty of these models, the lack of an inverse-Compton signature
from the galactic center means that only shallow density profiles can
describe the galactic dark matter density distribution
\cite{Bertone:2008xr,Cirelli:2008pk,Bergstrom:2008ag}. (For a review
of the dark matter modelling of these effects, see
\cite{Bergstrom:2009ib}. For a recent treatment, showing still viable
models, see \cite{Finkbeiner:2010sm}.)

A more standard explanation of the PAMELA/Fermi excess would be
pulsars or other supernova remnants (e.g.,
\cite{Hooper:2008kg,Ahlers:2009ae}). In that case, dark matter would
more likely be explained by the standard scenario of a WIMP, for
example the thoroughly studied lightest supersymmetric neutralino
(for reviews, see
\cite{Jungman:1995df,Bergstrom:2000pn,Bertone:2004pz}). We note that
the soon to be launched AMS-02 experiment on the International Space
Station \cite{ams02} may gave interesting clues to the origin of the
antimatter excess.

\subsubsection{Leptophilic models}
\label{sect:LP}
The results of PAMELA \cite{Adriani:2008zr} and Fermi-LAT
\cite{Abdo:2009zk} (and to some extent ATIC
\cite{2008Natur.456..362C}, although the bump seen there at 600 GeV
has not been reproduced by other experiments) has caused intense
activity in the theoretical dark matter community. At first, there was
skepticism whether such a large signal could be compatible with a
WIMP. However, the possibility of an enhanced annihilation rate due to
dark matter halo substructure has been realized for a long time
\cite{1993ApJ...411..439S,Bergstrom:1998zs,Moore:1999nt}, although it
is difficult to achieve the boost factor of the order of a few
hundred to a thousand in the solar neighborhood, as substructure
according to numerical simulations is most important in the outer
portions of the halo, due to tidal stripping of subhaloes in the inner
part. Another potentially very important effect which may explain the
large boost had been found a few years earlier, namely the Sommerfeld
enhancement. This effect, was computed for electromagnetism by Arnold
Sommerfeld may years ago \cite{sommerfeld} and recently rediscovered
in the quantum field theory of dark matter
\cite{2005PhRvD..71f3528H,2007NuPhB.787..152C,2009PhRvD..79a5014A}.
In quantum mechanics describing electron scattering and electron
positron annihilation, it is caused by the distortion of the plane
wave describing the relative motion of the annihilating particle pair,
due to the near formation of a bound state caused by photon
exchange. In the ladder approximation for QED, one reproduces the
Sommerfeld effect, and the square of the relative wave function at the
origin (which enters into the probability for the short-distance
process of annihilation) is increased by the factor
\cite{2009PhRvD..79a5014A}
$$
S_{QED}=\frac{|\psi(0)|^2}{|\psi^{(0)}(0)|^2}=|\frac{\frac{\pi\alpha}{v}}{1-e^{-\frac{\pi\alpha}{v}}}|,
$$ with $\alpha =e^2/\hbar c$ the fine-structure constant, and $v$ the
relative velocity. This amounts to $S_{QED}=\pi\alpha/v$ for small
velocities. For a Yukawa-like particle of mass $m_\phi$, mediating a
weak attractive force with coupling constant $\alpha_Y$ between dark
matter particles $\chi$ with mass $m_\chi$, the small-velocity limit
of the enhancement becomes instead
$$
S_Y\sim\frac{\alpha_Y m_\chi}{m_\phi}.
$$ In the general case, with mediators that may also excite virtual
charged particles in the dark matter sector, one has to solve
numerically a coupled system of differential equations using
appropriate boundary conditions
\cite{2005PhRvD..71f3528H,2007NuPhB.787..152C,2009PhRvD..79a5014A}. In
some cases, the enhancement factor $S$ can be as high as several
hundred to a few thousand, depending however on the exact parameters of
the theory. The effect is usually strongly velocity-dependent,
depending on velocity as $1/v$ or, in the fine-tuning case of being
very near resonance, as $1/v^2$. This means that in a virialised
system (such a a galaxy cluster) with large velocity dispersion
$v_{\rm cl}$ the Sommerfeld enhancement will be smaller than the one
expected in a single galaxy such as the Milky Way, roughly by a factor
$v_{\rm MW}/v_{\rm cl}$.

As leptophilic models, apart from being slightly contrived from the
particle physics point of view, are also rather limited by several
sets of astrophysical data, we take advantage of the recent reanalysis
\cite{Finkbeiner:2010sm} to define a benchmark leptophilic model that
is still viable.  It is found that the bounds from WMAP5 approximately
imply a Sommerfeld boost factor $S$ which has to satisfy $S(v\to
0)\lesssim 200/f\ (m_Y/1$ GeV$)$, where $f$ is the fraction of energy
from dark matter annihilation that ionizes the intergalactic medium,
$f\sim 0.7$ for annihilation to electrons and a factor a a few smaller
for all other standard model final states.

@@

To connect easily to our previous work \cite{2009PhRvL.103r1302P}, we
choose as our benchmark leptophilic model one with mass $1.6$ TeV and
Sommerfeld boost $S=300$.

\subsubsection{Supersymmetric dark matter}
The most studied models for dark matter are supersymmetric ones,
especially models where the lightest supersymmetric particle, in most
models the lightest neutralino, is stable. The stability is assured in
viable SUSY models due to a discrete symmetry, $R$-parity. This
symmetry is needed from the particle physics point of view to avoid
fast proton decay, and automatically makes the lightest supersymmetric
particle stable, and therefore a good candidate for dark matter. In
addition, as the coupling to ordinary matter is given by gauge
couplings, the neutralino is automatically a WIMP
candidate. Unfortunately, the breaking of supersymmetry (which has to
be there since otherwise, e.g., there would exist a scalar electron
with the same mass as the electron) means that a large number of
essentially free parameters enter actual calculations.  The neutralino
is then a linear combination of the supersymmetric partners of the
U(1) gauge boson, one of the SU(2) gauge bosons, and the two Higgs
doublets in the Minimal Supersymmetric extension of the Standard Model
(the MSSM) with mixing coefficients which depend on the SUSY breaking
parameters \cite{1984NuPhB.238..453E}. The parameter space is very
large, too large in general to scan efficiently even with the most
powerful computers. Therefore, often simplifying assumptions are made
to limit the number of free parameters. Even then, a large number of
viable models is found which give a relic density consistent with the
WMAP data. This means that they are indeed very good templates for
WIMP dark matter, something that explains their popularity besides the
subjective statement that few more plausible models for dark matter
have been put forward so far.

@@

@@

The photon spectrum resulting from FSR is universal with only a
weak dependence of the underlying particle physics model. The photon
yield from this process is given by (see
e.g. \cite{2008JHEP...01..049B})
\begin{equation}
\frac{\dd N_{X \bar{X}}}{\dd x} \approx \frac{\alpha Q_X^2}{\pi}
\mathcal{F}_X(x) \log\left[\frac{4 m_\chi^2\left(1-x\right)}{m_X^2}\right]\,.
\end{equation}
Here, the normalized photon energy $x=\eg/m_\chi c^2$, 
$\alpha =e^2/\hbar c$ is the fine-structure constant, 
$Q_X^2$ and $m_X$ the
charge and mass of the particle $X$, respectively. The function
$\mathcal{F}_X(x)$ depends on the spin of the final state and is given
by
\begin{equation}
\mathcal{F}_\rmn{fermion}(x) = \frac{1+\left(1-x\right)^2}{x}\,
\end{equation}
for fermions.

The differential energy spectrum for the IB process is more
complicated (see \cite{1989PhLB..225..372B,2008JHEP...01..049B}), but
can be computed using {\sc DarkSUSY}. The differential photon energy
spectra for all our benchmark models are shown in
Fig.~\ref{fig:bmspectra}.  The peaking at high photon energy and the
resulting bump caused by inner bremsstrahlung is clearly seen for
benchmark models $I´$, $J´$, and $J^*$.

\subsection{Astrophysics}
\label{sect:AP}

The differential photon flux within a given solid angle $\Delta
\Omega$ along a line-of-sight (los) is given by
\begin{equation}
\label{eq:dflux}
\frac{\dd \fg}{\dd \eg} \equiv \frac{\dd^3 \ngamma}{\dd A \,\dd t\, \dd
  \egamma} = \frac{1}{2}\int_{\Delta\Omega} \dd\psi \sin\psi \frac{\dd \sg}{\dd \eg}(\psi,\eg)\,,
\end{equation}
where
\begin{equation}
\label{eq:dtildeflux}
\frac{\dd \sg}{\dd \eg}(\psi,>\eg) = \int_{\Delta\Omega} \dd\Omega \int_{\rm los}
\dd l\, q_{\rm sum}\left(\eg,r\right)\Lambda(\theta)\,,
\end{equation}
and $\sg(\psi, >\eg)$ denotes the surface brightness above the photon
energy $\eg$.  The integration along the line-of-sight $l$, in the
direction $\psi$ that the detector is pointing, is parameterized such
that the radius of the source $r=\sqrt{l^2+D_\clu^2-2 D_\clu
  l\cos\Psi}$, where $D_\clu$ is the distance to the source from the
galactic center and
$\cos\Psi\equiv\cos\theta\cos\psi-\cos\varphi\sin\theta\sin\psi$. The
angular integration $\dd \Omega= \sin\theta\dd \theta \,\dd \varphi$
is performed over a cone centered around $\psi$ and the opening angle
$\Delta \Omega$ is typically taken to be a few times the point spread
function (PSF) $\psf$. The limited angular resolution results in a
probability that a photon coming from a direction $\psi$' is instead
reconstructed to a direction $\psi$, where the underlying probability
distribution follow a Gaussian:
\begin{equation}
\Lambda(\theta)=\frac{1}{2\pi\psf^2}
\,\rmn{exp}\left[-\frac{\theta^2}{2\psf^2}\right]\,,
\quad \rmn{where}\quad \theta=\psi'-\psi \,.
\end{equation}
We denote the total differential source function by $q_{\rm
  sum}(\eg,r)$, where we include contributions from five main
processes; leptophilic DM annihilating either indirectly through
$\mu^+/\mu^-$ to $\rmn{e}^+/\rmn{e}^-$ or directly to
$\rmn{e}^+/\rmn{e}^-$ pairs that inverse Compton upscatter background
photons (LP-IC), leptophilic DM emitting final state radiation
(LP-FSR), supersymmetric DM benchmark models where annihilating
neutralinos generate $\rmn{e}^+/\rmn{e}^-$ pairs that upscatter
background photons (BM-IC) and emit a continuum as well as final state
radiation (BM-Cont), and CR proton induced $\pi^0$ that decays into
gamma-rays (CR-$\pi^0$). The source function is given by
\begin{equation}
q_\rmn{sum} (\eg,r) = \qCR(\eg,r)+
\sum_i \,q_{\rmn{sm},i}(\eg,r)\,\B_{\rmn{tot},i}(\sigv,r)
\end{equation}
the differential CR to gamma-ray source function is denoted by
$\qCR(\eg,r)$ (see Sect.~\ref{sect:CRs} and \cite{2010MNRAS.409..449P}
for further details). The subscript $i$ runs over the gamma-ray
producing DM channels and the total differential boost factor for DM
is given by:
\begin{eqnarray}

>>>>

\author{Anders Pinzke$^{1}$} \email{apinzke@physics.ucsb.edu} 
\author{Christoph Pfrommer$^{2}$}\email{christoph.pfrommer@h-its.org}
\author{Lars Bergstr\"om$^{3}$}\email{lbe@fysik.su.se}

@@

\begin{abstract}
... to be written...  if $\mlim \approx 10^{-6}\,\msun$ then the boost
  from SFE is smaller than about a factor 5 in the MW and 2 in Fornax, or
  alternatively if SFE correct then $\mlim > 10^2\,\msun$.
\end{abstract}

@@

3. Galaxy clusters. This possibility has been less studied, however we
noted in a previous Letter \cite{2009PhRvL.103r1302P} that there are
certain advantages that work in favour of this possible target for
gamma-ray detection of dark matter annihilation.  Galaxy clusters
constitute the most massive objects in our Universe that are forming
today. This causes their DM subhalo mass function to be less affected
by tidal stripping compared to galaxy sized halos that formed long
ago.  The annihilation luminosities of the DM halo component for
e.g. the Virgo cluster and the Draco dwarf scales in a way (see
\cite{2009PhRvL.103r1302P}) that the ratio of gamma-ray luminosities
from the smooth components is around 4, in favour of Virgo. In
addition, there may be a further enhancement due to substructure,
which to a large extent should be unaffected by tidal disruption, at
least in the outer regions.  According to a recent estimate
\cite{2011MNRAS.410.2309G}, more massive haloes tend to have a larger
mass fraction in subhalos.  For example, cluster size haloes typically
have 7.5 per cent of the mass within $r_{200}$ in substructures of
fractional mass larger than $10^{-5}$, which is 25 per cent higher
than galactic haloes \footnote{All halo masses and length scales in
  this paper are scaled to the currently favored value of Hubble's
  constant, $H_0 = 70\, \rmn{km~s}^{-1}\,\rmn{Mpc}^{-1}$. We define the
  virial mass $\mvir$ and virial radius $\rvir$ as the mass and radius
  of a sphere enclosing a mean density that is 200 times the critical
  density of the Universe $\rho_{\rmn{cr}}$.}.

@@

This yields Fornax ($\mvir=10^{14}\,\msun$) and Virgo
($\mvir=2.1\times10^{14}\,\msun$) \del{\cite{}} as the prime targets for DM
observations, Perseus ($\mvir=7.7\times10^{14}\,\msun$) for CR induced emission
and we additionally decide in favor of the well studied cluster Coma
($\mvir=1.4\times10^{15}\,\msun$) for comparison. While studying these four
clusters in great detail, we complement the analysis by calculating the expected
gamma-ray fluxes of a nearby X-ray flux complete sample of galaxy clusters
(HIFLUGCS, \citealt{2002ApJ...567..716R})

@@

This work studies the prospects for detecting extended gamma-ray
emission from galaxy clusters induced by both DM as well as CR
protons. In Section 2 we discuss the theory of DM and gamma-rays. In
particular, we focus one the leptophilic and supersymmetric DM models
used in this work, and outline the framework for estimating the
gamma-ray emission from various radiative processes. In Section 3
we calculate the gamma-ray fluxes and spectral distributions from four
clusters and compare it to upper limits set by Fermi-LAT. For the same
clusters we derive the gamma-ray surface brightness profiles in
Section 4. To extend the analysis to more clusters and find
promising targets we estimate the fluxes from DM and CRs of all
clusters in the extended HIFLUGCS sample in Section 5. In Section
6 we conclude and discuss our results.

% --- section:  --- %
\section{Theory}
\label{sect:theory}
We start the section by discussing two different, but well motivated DM models;
leptophilic and supersymmetric DM. We then present the framework that is used to
calculate the gamma-ray emission for these DM models using an Einasto DM density
profile and the expected enhancement from DM substructures. This is followed by
an outline of the framework of inverse Compton emission where we take into
account photon fields of the cosmic microwave background (CMB), dust, and
starlight. We end by summarizing our formalism of calculating the CR-induced
pion decay gamma-ray emission which is thought to dominate the astrophysical
gamma-ray signal.

\subsection{Detecting Particle Dark Matter}
\label{sect:PF}
Besides the intrinsic interest in the gamma-ray flux from galaxy
clusters generated by conventional hadronic and electromagnetic
processes - which should be close to observability with the Fermi-LAT
data \cite{1997ApJ...487..529B,2007A&A...473...41E,2010MNRAS.409..449P}
- the possible contribution from WIMP dark matter is of greatest
interest. A WIMP (Weakly Interactive Massive Particle) which fulfils
the WMAP bounds on the relic density of cold dark matter (CDM) of
\cite{Komatsu:2010fb}
$$\Omega_{CDM}h^2=0.112\pm 0.0056,$$ will in many cases naturally give
a gamma-yield which may be observable. In addition, there are
possible enhancement effects known, such as the astrophysical boost
from dense substructure of dark matter halos, or the particle physics
boost from the Sommerfeld effect, which may increase the chances of
detection further.

There are three methods for detecting dark matter candidates which are
presently employed: First, at particle physics accelerators such as
CERN's LHC, one is now entering into a new energy regime which may
allow the production of the heavy, electrically neutral and long-lived
particles which may constitute dark matter. From these experiments one
may get the first glimpse of the mass scale beyond the standard model
where dark matter may reside. However, to really show that any of the
hypothetical, new particles created at the LHC is actually the dark
matter, one has to rely on the two other methods available for the
dark matter search, namely direct and indirect detection. Direct
detection methods, which are presently evolving rapidly use the feeble
interaction between dark matter particles of the galactic halo
and nuclei such as Germanium, Sodium, Iodine, Argon or Xenon to infer
the scattering cross section and a rough estimate of the mass of the
dark matter particle (for the currently most sensitive search, see
\cite{Aprile:2010um}). A characteristic of this method is that it
basically only depends on the local dark matter density (which is
rather well determined to be around $0.4$ GeV/cm$^3$) and the basic
cross section for dark matter - nucleus scattering. A disadvantage is
that it does benefit from the two enhancement mechanisms mentioned
above, boost from substructure and/or the Sommerfeld effect. Also, it
cannot be excluded - although it seems at present improbable - that
the local halo structure is such that the solar system happens to be
in an underdense region.

@@

There is also a possibility to search for dark matter annihilation indirectly in
antimatter channels, like antiprotons or positrons. Here diffusion is, however,
very efficient which means that the directional signature which is so important
for gamma-rays, is completely erased. This method has, however, recently been
much in focus due to the surprising rise with energy of the positron-to-total
lepton ratio measured by the PAMELA satellite experiment \cite{Adriani:2008zr}
and the ``bump'' in the total electron plus positron spectrum detected by Fermi
\cite{Abdo:2009zk}. The viable particle physics models describing these results
contain some non-trivial, or even unnatural elements, however. First of all,
annihilation to quarks has to be very much suppressed, due to the
non-observation of an enhanced antiproton flux by PAMELA
\cite{Adriani:2010rc}. This means that only so-called {\em leptophilic} models,
where the annihilation only goes to leptons have to be considered. Here, one has
to have a sizeable fraction going to muons and tau leptons, as the shape of the
spectrum disfavours direct annihilation into electrons and positrons. To get
agreement with the measured absolute rates, one also has to have a large boost
factor (of the order of at least several hundred), which is difficult to achieve
with substructure alone. An additional enhancement of the annihilation rate has
to be invoked, for example given by the {\em Sommerfeld effect}. This basically
means that the usual connection between annihilation rate at decoupling (when
relative velocities $v/c$ were of order $0.2 - 0.3$), which gives the relic
density, and present-day annihilation in the halo (with velocities $v/c\sim
10^{-3}$) is broken. This would come about because the dark matter sector
contains additional light particles which mediates an interaction with almost
bound states which enhance the wave function at the relative origin, and
therefore the annihilation rate (for a pedagogical review of viable scenarios,
see \cite{ArkaniHamed:2008qn}). As a final difficulty of these models, the lack
of an inverse-Compton signature from the galactic center means that only shallow
density profiles can describe the galactic dark matter density distribution
\cite{Bertone:2008xr,Cirelli:2008pk,Bergstrom:2008ag}. (For a review of the dark
matter modelling of these effects, see \cite{Bergstrom:2009ib}. For a recent
treatment, showing still viable models, see \cite{Finkbeiner:2010sm}.)

A more standard explanation of the PAMELA/Fermi excess would be
pulsars or other supernova remnants (e.g.,
\cite{Hooper:2008kg,Ahlers:2009ae}). In that case, dark matter would
more likely be explained by the standard scenario of a WIMP, for
example the thoroughly studied lightest supersymmetric neutralino
(for reviews, see
\cite{Jungman:1995df,Bergstrom:2000pn,Bertone:2004pz}). We note that
the soon to be launched AMS-02 experiment on the International Space
Station \cite{ams02} may give interesting clues to the origin of the
antimatter excess.

\subsubsection{Leptophilic models}
\label{sect:LP}
The results of PAMELA \cite{Adriani:2008zr} and Fermi-LAT
\cite{Abdo:2009zk} (and to some extent ATIC
\cite{2008Natur.456..362C}, although the bump seen there at 600 GeV
has not been reproduced by other experiments) has caused intense
activity in the theoretical dark matter community. At first, there was
skepticism whether such a large signal could be compatible with a
WIMP. However, the possibility of an enhanced annihilation rate due to
dark matter halo substructure has been realized for a long time
\cite{1993ApJ...411..439S,Bergstrom:1998zs,Moore:1999nt}, although it
is difficult to achieve the boost factor of the order of a few
hundred to a thousand in the solar neighborhood, as substructure
according to numerical simulations is most important in the outer
portions of the halo, due to tidal stripping of subhaloes in the inner
part. Another potentially very important effect which may explain the
large boost had been found a few years earlier, namely the Sommerfeld
enhancement. This effect, was computed for electromagnetism by Arnold
Sommerfeld many years ago \cite{sommerfeld} and recently rediscovered
in the quantum field theory of dark matter
\cite{2005PhRvD..71f3528H,2007NuPhB.787..152C,2009PhRvD..79a5014A}.
In quantum mechanics describing electron scattering and electron
positron annihilation, it is caused by the distortion of the plane
wave describing the relative motion of the annihilating particle pair,
due to the near formation of a bound state caused by photon
exchange. In the ladder approximation for QED, one reproduces the
Sommerfeld effect, and the square of the relative wave function at the
origin (which enters into the probability for the short-distance
process of annihilation) is increased by the factor
\cite{2009PhRvD..79a5014A}
$$
S_{QED}=\frac{|\psi(0)|^2}{|\psi^{(0)}(0)|^2}=
\left|\frac{\frac{\pi\alpha}{v}}{1-e^{-\frac{\pi\alpha}{v}}}\right|,
$$ 
with $\alpha =e^2/\hbar c$ the fine-structure constant, and $v$ the relative
velocity. This amounts to $S_{QED}=\pi\alpha/v$ for small velocities. For a
Yukawa-like particle of mass $m_\phi$, mediating a weak attractive force with
coupling constant $\alpha_Y$ between dark matter particles $\chi$ with mass
$m_\chi$, the small-velocity limit of the enhancement becomes instead
$$
S_Y\sim\frac{\alpha_Y m_\chi}{m_\phi}.
$$ In the general case, with mediators that may also excite virtual
charged particles in the dark matter sector, one has to solve
numerically a coupled system of differential equations using
appropriate boundary conditions
\cite{2005PhRvD..71f3528H,2007NuPhB.787..152C,2009PhRvD..79a5014A}. In
some cases, the enhancement factor $S$ can be as high as several
hundred to a few thousand, depending however on the exact parameters of
the theory. The effect is usually strongly velocity-dependent,
depending on velocity as $1/v$ or, in the fine-tuning case of being
very near resonance, as $1/v^2$. This means that in a virialised
system (such as a galaxy cluster) with large velocity dispersion
$v_{\rm cl}$ the Sommerfeld enhancement will be smaller than the one
expected in a single galaxy such as the Milky Way, roughly by a factor
$v_{\rm MW}/v_{\rm cl}$.

As leptophilic models, apart from being slightly contrived from the
particle physics point of view, are also rather limited by several
sets of astrophysical data, we take advantage of the recent reanalysis
\cite{Finkbeiner:2010sm} to define a benchmark leptophilic model that
is still viable.  It is found that the bounds from WMAP5 approximately
imply a Sommerfeld boost factor $S$ which has to satisfy $S(v\to
0)\lesssim 200/f\ (m_Y/1$ GeV$)$, where $f$ is the fraction of energy
from dark matter annihilation that ionizes the intergalactic medium,
$f\sim 0.7$ for annihilation to electrons and a factor of a few smaller
for all other standard model final states.

@@

To connect easily to our previous work \cite{2009PhRvL.103r1302P}, we choose as
our benchmark leptophilic model one with mass $1.6$ TeV and Sommerfeld boost
$S=300$ for a Milky Way halo and scale this appropriately down for cluster halos, using

{\bf Anders: insert Sommerfeld boost formula as a function of halo mass}

Clusters have a larger velocity dispersion and an accordingly smaller Sommerfeld
boost factor.

\subsubsection{Supersymmetric dark matter}
The most studied models for dark matter are supersymmetric ones, especially
models where the lightest supersymmetric particle, in most models the lightest
neutralino, is stable. The stability is assured in viable SUSY models due to a
discrete symmetry, $R$-parity. This symmetry is needed from the particle physics
point of view to avoid fast proton decay, and automatically makes the lightest
supersymmetric particle stable, and therefore a good candidate for dark
matter. In addition, as the coupling to ordinary matter is given by gauge
couplings, the neutralino is automatically a WIMP candidate. Unfortunately, the
breaking of supersymmetry (which has to be there since otherwise, e.g., there
would exist a scalar electron with the same mass as the electron) means that a
large number of essentially free parameters enter actual calculations.  The
neutralino is then a linear combination of the supersymmetric partners of the
U(1) gauge boson, $\tilde B$, one of the SU(2) gauge bosons, $\tilde W_3$, and
the two Higgs doublets in the Minimal Supersymmetric extension of the Standard
Model (the MSSM), $\tilde H_1^0$ and $\tilde H_2^0$, with mixing coefficients
which depend on the SUSY breaking parameters \cite{1984NuPhB.238..453E}. The
parameter space is very large, too large in general to scan efficiently even
with the most powerful computers. Therefore, often simplifying assumptions are
made to limit the number of free parameters. Even then, a large number of viable
models is found which give a relic density consistent with the WMAP data. This
means that they are indeed very good templates for WIMP dark matter, something
that explains their popularity besides the subjective statement that few more
plausible models for dark matter have been put forward so far.

@@

{\bf Anders: add DM particle mass and cross section to all these 4 models}

@@

The photon spectrum resulting from FSR is universal with only a
weak dependence of the underlying particle physics model. The photon
yield from this process is given by (see
e.g. \cite{2008JHEP...01..049B})
\begin{equation}
\frac{\dd N_{X \bar{X}}}{\dd x} \approx \frac{\alpha Q_X^2}{\pi}
\mathcal{F}_X(x) \log\left[\frac{4 m_\chi^2\left(1-x\right)}{m_X^2}\right]\,.
\end{equation}
Here, the normalized photon energy $x=\eg/m_\chi c^2$, 
$\alpha =e^2/\hbar c$ is the fine-structure constant, 
$Q_X^2$ and $m_X$ the
charge and mass of the final state particle $X$, respectively. The function
$\mathcal{F}_X(x)$ depends on the spin of the final state and is given
by
\begin{equation}
\mathcal{F}_\rmn{fermion}(x) = \frac{1+\left(1-x\right)^2}{x}\,
\end{equation}
for fermions.

The differential energy spectrum for the IB process is more
complicated (see \cite{1989PhLB..225..372B,2008JHEP...01..049B}), but
can be computed using {\sc DarkSUSY}. The differential photon energy
spectra for all our benchmark models are shown in Fig.~\ref{fig:q_DM}.
The peaking at high photon energy and the resulting bump caused by
inner bremsstrahlung is clearly seen for benchmark models $I´$, $J´$,
and $J^*$. Despite the different particle masses and cross sections,
the spectrum of the continuums emission shows a remarkably similar
shape which suggests that a simple rescaling of particle masses
(responsible for the exponential cutoff in the spectrum) and cross
sections (spectral amplitude) should yield a roughly scale-invariant
contiuums emission spectrum except for the presence of the final state
emission feature which depends on the details of the specific decay
channels. The right panel of Fig.~\ref{fig:q_DM} shows the cumuative
number of leptons (electrons and positrons) above a given energy for
our different DM models which is of interested for the high-energy
inverse Compton emission of those leptons.  Only leptophilic (LP)
models have energetic enough electrons such that the IC emission is
powerful enough to either be constrained or detected at GeV energies
and higher.

\begin{figure*}
\begin{minipage}{2.0\columnwidth}
 \includegraphics[width=0.49\columnwidth]{figures/fit.ds.flux.eps}
 \includegraphics[width=0.49\columnwidth]{figures/fit.epflux.int.eps}
 \caption{\it Source functions for different DM models. Crosses show
   the simulated data from DarkSUSY and the solid lines show the fit
   to the data using Eq.... LEFT PANEL: normalized differential
   continuum spectra for four different DM benchmark (BM) models;
   $\Ip$ model (red), $\Jp$ (orange), $\Kp$ (green), and $\Js$
   (blue). RIGHT PANEL: number of electron and positron per DM
   annihilation above the electron energy $\ee$ for different dark
   matter models; $\Ip$ BM model (red), $\Jp$ BM model (orange), $\Kp$
   BM model (green), and $\Js$ BM model (blue), leptophilic (LP) DM
   annihilating into electrons and positrons (purple), and LP DM
   annihilating into muons (black) without considering Sommerfeld
   boosts.}
 \label{fig:q_DM}
\end{minipage}
\end{figure*}

\subsection{Astrophysical modeling of DM induced emission}
\label{sect:AP}

We now turn to the detailed modeling of the surface brightness
profiles of DM annihilation emission and discuss the DM profiles for
the smooth distribution and substructures.

\subsubsection{General equations}
